Scientific Reasoning: \\ Assessment of Multimodal Generative LLMs

### 2025-02-08 docs/paper/main.tex
documentation

@@

\begin{icmlauthorlist}
\icmlauthor{Florian Dreyer (3667877)}{equal,cs}
\icmlauthor{Ekaterina Kolos (3688474)}{equal,ims}
\icmlauthor{Daria Matiash (3668740)}{equal,ims}
\end{icmlauthorlist}

\icmlaffiliation{cs}{Institut fuer KI}
\icmlaffiliation{ims}{Institut fuer Maschinelle Sprachverarbeitung}

\icmlcorrespondingauthor{Florian Dreyer}{c.vvvvv@googol.com}
%\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Scientific Reasoning, VQA}

@@

\begin{abstract}
This project assesses the capabilities of pre-trained multimodal LLMs to perform scientific reasoning tasks on multimodal data. We further explore how prompt tuning (soft prompting) and prompt engineering techniques can improve performance. We then attempt to distill knowledge to a small LLM.
\end{abstract}

@@

\textbf{SECTION NEEDS UPDATE}
Scientific machine reasoning is the application of AI techniques to simulate scientific reasoning processes, such as  data interpretation, hypothesis generation, and causal inference. It facilitates scientific inquiry and discovery by enabling machines to independently analyze scientific data, generate hypotheses, and make predictions.
For our project, we intend to compare different approaches around generative large language models to generate solutions for school science problems. On an existing benchmark with human solutions, we are planning to compare: a pure LLM with zero-shot and few-shot prompting (and possibly some more fine-grained prompting techniques), an agentic approach with ability to query external resources like textbooks in physics, maths etc, and soft-prompting techniques. We then plan to attempt knowledge distillation to a smaller LLM.

\section{Motivation}
Recent developments in foundation models have shown remarkable performance on challenges that call for contextual awareness and scientific reasoning. However, these models are often resource-intensive, which limitates their scalability and accessibility for broader applications. Through the transfer of reasoning capacities from large models to smaller, more efficient ones, this study focuses on the possibility of knowledge distillation to address these issues. By applying this approach to the Science QA datase, we want to close the gap between performance and usefulness by enabling lightweight models to handle challenging reasoning tasks with high accuracy and robustness.
\label{motivation}
\textbf{TODO}

>>>>

%6 pages + references

@@

\begin{icmlauthorlist}
\icmlauthor{Florian Dreyer (3667877)}{equal,us}
\icmlauthor{Ekaterina Kolos (3688474)}{equal,us}
\icmlauthor{Daria Matiash (3668740)}{equal,us}
\end{icmlauthorlist}

\icmlaffiliation{us}{University of Stuttgart}
%\icmlaffiliation{ims}{Institut fuer Maschinelle Sprachverarbeitung}

\icmlcorrespondingauthor{Florian Dreyer}{st182762@stud.uni-stuttgart.de}
\icmlcorrespondingauthor{Ekaterina Kolos}{st186032@stud.uni-stuttgart.de}
\icmlcorrespondingauthor{Daria Matiash}{st185745@stud.uni-stuttgart.de}
%\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Scientific Reasoning, Prefix Tuning, Knowledge Distillation}

@@

\begin{abstract}
%Note: we are doing scientific QA and not really scientific reasoning in the end. How do we frame it? -> Added QA to multimodal data, what do you think? -- Good, thanks!
This project assesses the capabilities of pre-trained multimodal LLMs to perform scientific reasoning tasks on multimodal Question Answering (QA) data. We further explore how Prefix Tuning and Low-Rank Adaptation (LoRA) can improve performance of smaller LLMs. We then attempt to distill knowledge to a smaller LLM using Prefix Tuning and LoRA.
\end{abstract}

@@

Scientific machine reasoning is the application of AI techniques to simulate scientific reasoning processes, such as data interpretation, hypothesis generation, and causal inference. 
%It facilitates scientific inquiry and discovery by enabling machines to independently analyze scientific data, generate hypotheses, and make predictions.

Recent developments in foundation models, such as o1/o3 from OpenAI \cite{OpenAI2024o1} or R1 from DeepSeek \cite{deepseek-r1}, have shown remarkable performance in challenges that require contextual awareness and reasoning.
However, these models are often resource consuming, which limits their scalability and accessibility for broader applications.

This leads us to our research questions:

\paragraph{RQ1}
\label{def:rq1}
How can large multimodal LLMs deal with multimodal scientific reasoning?
%We in fact answer three questions: 
% 1) how well they perform in scientific question answering (QTCH), 
% 2) whether they are capable of reasoning (QTCHL) 
% 3) how well they use in-context information (=solution) to produce expected useful generation

\paragraph{RQ2}
\label{def:rq2}
How do Prefix Tuning and LoRA affect the reasoning capabilities of smaller pre-trained models?
% This is only scientific QA, without reasoning!

\paragraph{RQ3}
\label{def:rq3}
How good is knowledge distillation with adapter methods compared to training on manually annotated data?

We first want to evaluate how six large front-tier LLMs perform on the \textsc{ScienceQA} dataset \cite{lu2022learn} using several metrics. Following that we evaluate how well Prefix Tuning \cite{li2021prefix} and LoRA \cite{hu2021lora} perform for fine-tuning two smaller LLMs on this dataset. Last, we perform knowledge distillation using Prefix Tuning and LoRA and compare how the distilled models compare to the previously fine-tuned models.
